---
title: "Validating fruitArchitecture 0.2.0.9001"
author: "Franklin T. Johnson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Validating fruitArchitecture 0.2.0.9001}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  error = FALSE,
  warning = TRUE,
  message = TRUE
)

assert_true <- function(condition, message) {
  if (!isTRUE(condition)) {
    stop(message, call. = FALSE)
  }
  invisible(TRUE)
}

assert_identical <- function(object, expected, message) {
  if (!identical(object, expected)) {
    stop(
      paste0(
        message,
        "\nObserved: ", paste(object, collapse = ", "),
        "\nExpected: ", paste(expected, collapse = ", ")
      ),
      call. = FALSE
    )
  }
  invisible(TRUE)
}
```

# Purpose

This vignette is a release-validation document for `fruitArchitecture`
version `0.2.0.9000`. It tests the frozen legacy PHMIES definition, the
new Paper 5 Broad6 definition, the complete 15-interface universe, and the
frozen Core10-to-Broad6 assignment registry.

The validation is intentionally independent of raw RNA-seq files. It can be
rebuilt during `R CMD check` and used before creating the next GitHub
prerelease tag.

# Load the installed package

```{r load-package}
library(fruitArchitecture)

package_version <- as.character(
  utils::packageVersion("fruitArchitecture")
)

package_version
```

```{r validate-version}
package_version <- as.character(
  utils::packageVersion("fruitArchitecture")
)

if (
  utils::compareVersion(
    package_version,
    "0.2.0.9000"
  ) < 0L
) {
  stop(
    paste0(
      "fruitArchitecture 0.2.0.9000 or later is required to run 'broad6' definition in the fruitArchitecture call. ",
      "Observed: ",
      package_version
    ),
    call. = FALSE
  )
}
```

# Retrieve the internal frozen definitions

These validation calls use `getFromNamespace()` because the definition
constructors are internal package infrastructure rather than public user
functions.

```{r retrieve-internal-functions}
resolve_architecture <- getFromNamespace(
  ".resolve_architecture_definition",
  "fruitArchitecture"
)

broad6_interface_registry <- getFromNamespace(
  ".fa_broad6_interfaces",
  "fruitArchitecture"
)
```

# Validate the frozen PHMIES definition

The default architecture must remain the original PHMIES definition so that
analyses performed under `fruitArchitecture` v0.1 remain reproducible.

```{r validate-phmies}
legacy_definition <- resolve_architecture(NULL)

legacy_summary <- data.frame(
  field = c(
    "name",
    "version",
    "definition_status",
    "module_definition",
    "circuit_definition",
    "module_count",
    "required_interface_count"
  ),
  value = c(
    legacy_definition$name,
    legacy_definition$version,
    legacy_definition$definition_status,
    legacy_definition$module_definition,
    legacy_definition$circuit_definition,
    length(legacy_definition$modules),
    nrow(legacy_definition$interfaces)
  ),
  stringsAsFactors = FALSE
)

legacy_summary
```

```{r assert-phmies}
assert_identical(
  legacy_definition$name,
  "PHMIES",
  "The default architecture definition is no longer PHMIES."
)

assert_identical(
  legacy_definition$module_definition,
  "core10",
  "The PHMIES definition must resolve to the frozen Core10 system."
)

assert_identical(
  legacy_definition$circuit_definition,
  "none",
  "The legacy PHMIES definition must not activate a directed circuit."
)

assert_true(
  length(legacy_definition$modules) == 10L,
  "The frozen PHMIES definition must contain 10 modules."
)

assert_true(
  nrow(legacy_definition$interfaces) == 3L,
  "The frozen PHMIES definition must contain three required interfaces."
)
```

# Validate the Paper 5 Broad6 definition

```{r validate-paper5-definition}
paper5_definition <- resolve_architecture(
  "paper5_frozen"
)

paper5_summary <- data.frame(
  field = c(
    "name",
    "version",
    "definition_status",
    "module_definition",
    "circuit_definition",
    "module_count",
    "required_interface_count",
    "interface_universe_count"
  ),
  value = c(
    paper5_definition$name,
    paper5_definition$version,
    paper5_definition$definition_status,
    paper5_definition$module_definition,
    paper5_definition$circuit_definition,
    length(paper5_definition$modules),
    nrow(paper5_definition$interfaces),
    nrow(paper5_definition$interface_universe)
  ),
  stringsAsFactors = FALSE
)

paper5_summary
```

```{r assert-paper5-definition}
expected_broad6_modules <- c(
  "Circadian",
  "Hormone",
  "MAPK",
  "Information Exchange",
  "RNA regulation",
  "Proteostasis"
)

assert_identical(
  paper5_definition$name,
  "paper5_frozen",
  "The Paper 5 architecture name is incorrect."
)

assert_identical(
  paper5_definition$module_definition,
  "broad6",
  "The Paper 5 architecture must use Broad6."
)

assert_identical(
  paper5_definition$circuit_definition,
  "autocatalytic_ethylene_v1",
  "The Paper 5 definition must identify the frozen ethylene circuit."
)

assert_identical(
  paper5_definition$modules,
  expected_broad6_modules,
  "The frozen Broad6 module order or names changed."
)

assert_true(
  isFALSE(paper5_definition$extended_interfaces_affect_class),
  paste(
    "Extended Broad6 interfaces must not independently alter",
    "the frozen PHMIES Class I-IV call."
  )
)
```

# Validate the complete Broad6 interface universe

```{r build-interface-registry}
broad6_interfaces <- broad6_interface_registry()

broad6_interfaces
```

```{r assert-interface-registry}
assert_true(
  is.data.frame(broad6_interfaces),
  "The Broad6 interface registry must be a data frame."
)

assert_true(
  nrow(broad6_interfaces) == 15L,
  "Six Broad6 modules must generate exactly 15 pairwise interfaces."
)

assert_true(
  length(unique(broad6_interfaces$interface_id)) == 15L,
  "Broad6 interface identifiers must be unique."
)

required_interface_ids <- c(
  "Circadian--Hormone",
  "Circadian--Proteostasis",
  "Hormone--MAPK",
  "Hormone--Information Exchange",
  "Information Exchange--MAPK"
)

assert_true(
  all(required_interface_ids %in% broad6_interfaces$interface_id),
  "One or more required Broad6 interface identifiers are absent."
)

assert_true(
  all(
    paper5_definition$level3a_interface_ids %in%
      broad6_interfaces$interface_id
  ),
  paste(
    "Every Paper 5 Level 3A interface must be represented",
    "in the complete Broad6 interface universe."
  )
)
```

Canonical interface identifiers are orientation-independent. Therefore,
`MAPK` and `Information Exchange` are stored as
`Information Exchange--MAPK`, while the display label may retain the
biological reading order `MAPK-Information Exchange`.

# Validate the frozen Core10-to-Broad6 registry

The package must install the authoritative mapping table at:

```text
inst/extdata/module_registry_broad6_v1.csv
```

```{r locate-module-registry}
registry_path <- system.file(
  "extdata",
  "module_registry_broad6_v1.csv",
  package = "fruitArchitecture"
)

assert_true(
  nzchar(registry_path) && file.exists(registry_path),
  paste(
    "The frozen Broad6 registry was not installed.",
    "Expected inst/extdata/module_registry_broad6_v1.csv."
  )
)

module_registry <- utils::read.csv(
  registry_path,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

utils::head(module_registry)
```

```{r assert-module-registry}
required_registry_columns <- c(
  "source_definition",
  "source_module",
  "broad6_module",
  "assignment_status",
  "evidence_source",
  "definition_version",
  "notes"
)

assert_true(
  all(required_registry_columns %in% names(module_registry)),
  "The Broad6 registry is missing one or more required columns."
)

assert_true(
  nrow(module_registry) == 41L,
  paste(
    "The frozen registry must contain 41 rows:",
    "10 canonical Core10 assignments and 31 aliases."
  )
)

assert_true(
  sum(module_registry$source_definition == "core10_canonical") == 10L,
  "The registry must contain exactly 10 canonical Core10 assignments."
)

assert_true(
  sum(module_registry$source_definition == "core10_alias") == 31L,
  "The registry must contain exactly 31 frozen alias assignments."
)

assert_identical(
  sort(unique(module_registry$broad6_module)),
  sort(expected_broad6_modules),
  "The registry does not represent exactly the six frozen Broad6 modules."
)

assert_identical(
  unique(module_registry$definition_version),
  "broad6-v1.0",
  "The module registry definition version changed unexpectedly."
)

assert_true(
  all(
    module_registry$assignment_status %in%
      c("frozen_canonical", "frozen_alias")
  ),
  "The registry contains an unrecognized assignment status."
)

assert_true(
  all(nzchar(module_registry$evidence_source)),
  "Every registry row must contain an evidence source."
)

assert_true(
  all(nzchar(module_registry$notes)),
  "Every registry row must contain an explanatory note."
)
```

## Canonical mapping

```{r canonical-mapping}
canonical_mapping <- module_registry[
  module_registry$source_definition == "core10_canonical",
  c(
    "source_module",
    "broad6_module",
    "assignment_status",
    "definition_version"
  )
]

canonical_mapping
```

## Assignment counts by Broad6 module

```{r registry-counts}
assignment_counts <- as.data.frame(
  table(module_registry$broad6_module),
  stringsAsFactors = FALSE
)

names(assignment_counts) <- c(
  "broad6_module",
  "registry_row_count"
)

assignment_counts[
  order(assignment_counts$broad6_module),
]
```

# Inspect the public API

The legacy argument must remain available. The additional versioned arguments
can be added without changing the established positional arguments.

```{r inspect-api}
fruit_architecture_arguments <- names(
  formals(fruitArchitecture)
)

api_status <- data.frame(
  argument = c(
    "architecture_definition",
    "module_definition",
    "circuit_definition",
    "regulatory_overlay"
  ),
  implemented = c(
    "architecture_definition",
    "module_definition",
    "circuit_definition",
    "regulatory_overlay"
  ) %in% fruit_architecture_arguments,
  stringsAsFactors = FALSE
)

api_status
```

```{r assert-legacy-api}
assert_true(
  "architecture_definition" %in% fruit_architecture_arguments,
  paste(
    "fruitArchitecture() must retain the",
    "architecture_definition argument."
  )
)
```

The following check reports, but does not stop the vignette, while the new
public API is being connected to the underlying frozen definitions.

```{r report-new-api}
new_api_arguments <- c(
  "module_definition",
  "circuit_definition",
  "regulatory_overlay"
)

missing_new_api <- setdiff(
  new_api_arguments,
  fruit_architecture_arguments
)

if (length(missing_new_api) == 0L) {
  message(
    "The complete fruitArchitecture 0.2 public API is present."
  )
} else {
  message(
    "Definition infrastructure is valid. Public API arguments still to add: ",
    paste(missing_new_api, collapse = ", ")
  )
}
```

# Optional installed-example smoke test

This section looks for the original lightweight example files. It performs the
legacy analysis only when all three files are installed. A failed analysis is
reported with its original error message so that input-format changes can be
diagnosed.

```{r example-smoke-test}
example_paths <- c(
  counts = system.file(
    "extdata",
    "example_counts.csv",
    package = "fruitArchitecture"
  ),
  design = system.file(
    "extdata",
    "example_design.csv",
    package = "fruitArchitecture"
  ),
  annotation = system.file(
    "extdata",
    "example_annotation.csv",
    package = "fruitArchitecture"
  )
)

example_files_available <- all(
  nzchar(example_paths) & file.exists(example_paths)
)

if (!example_files_available) {
  message(
    "Installed example files were not all found; ",
    "the analysis smoke test was skipped."
  )
} else {
  example_counts_df <- utils::read.csv(
    example_paths[["counts"]],
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  example_design <- utils::read.csv(
    example_paths[["design"]],
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  example_annotation <- utils::read.csv(
    example_paths[["annotation"]],
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  # Most package count examples store gene identifiers in the first column.
  example_counts <- example_counts_df

  legacy_smoke_result <- tryCatch(
    fruitArchitecture(
      counts = example_counts,
      design = example_design,
      annotation = example_annotation,
      architecture_definition = "PHMIES"
    ),
    error = function(e) e
  )

  if (inherits(legacy_smoke_result, "error")) {
    message(
      "Legacy example analysis did not complete: ",
      conditionMessage(legacy_smoke_result)
    )
  } else {
    message(
      "Legacy example analysis completed successfully."
    )
    print(legacy_smoke_result)
  }
}
```

# Paper 5 public-call template

After `module_definition`, `circuit_definition`, and `regulatory_overlay` are
connected to `fruitArchitecture()`, the intended call is:

```{r paper5-call-template, eval=FALSE}
paper5_result <- fruitArchitecture(
  counts = counts,
  design = design,
  annotation = annotation,
  module_definition = "broad6",
  architecture_definition = "paper5_frozen",
  circuit_definition = "autocatalytic_ethylene_v1",
  regulatory_overlay = validated_regulatory_overlay
)
```

The annotation-derived architecture and experimentally augmented regulatory
architecture should remain distinct output objects. A directed circuit should
be called engaged only when all required nodes, curated edges, species, and
developmental-context conditions are supported.

# GSE62415 regression target

The GSE62415 apple-seed validation should later be added as a separate article
or example because rebuilding raw RNA-seq data is not appropriate during
routine package checks.

The frozen expected result under the legacy PHMIES rules is:

```{r gse62415-target}
gse62415_expected <- data.frame(
  metric = c(
    "validation_type",
    "architecture_definition",
    "module_definition",
    "architecture_class",
    "level3a_present",
    "level3b_count",
    "differentially_engaged_phmies_interface_genes"
  ),
  expected_value = c(
    "module-positive/interface-negative",
    "PHMIES",
    "core10",
    "I",
    "FALSE",
    "0",
    "0"
  ),
  stringsAsFactors = FALSE
)

gse62415_expected
```

# Validation result

Reaching this section means that all required hard assertions in the vignette
passed.

```{r validation-result}
validation_result <- data.frame(
  check = c(
    "Package version",
    "Legacy PHMIES definition",
    "Paper 5 Broad6 definition",
    "Broad6 interface count",
    "Level 3A interface consistency",
    "Frozen module registry",
    "Legacy public API"
  ),
  status = rep("PASS", 7L),
  stringsAsFactors = FALSE
)

validation_result
```

The package is ready for the next implementation stage when this vignette
builds successfully, `devtools::test()` passes, and `R CMD check` reports no
package-content notes.
