---
title: "Getting started with fruitArchitecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fruitArchitecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fruitArchitecture)
```

# Overview

`fruitArchitecture` separates differential-expression analysis from
architecture reconstruction. Users can begin with either a validated DEG table
or a raw RNA-seq count matrix.

# Installation into the user library

```{r installation, eval=FALSE}
user_library <- .libPaths()[1]
dir.create(user_library, recursive = TRUE, showWarnings = FALSE)

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes", lib = user_library)
}

remotes::install_local(
  "C:/path/to/fruitArchitecture",
  lib = user_library,
  dependencies = c("Depends", "Imports", "LinkingTo"),
  upgrade = "never",
  force = TRUE
)
```

Install DESeq2 only when raw counts will be analyzed:

```{r deseq2-installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("DESeq2", ask = FALSE, update = FALSE)
```

# Included example files

```{r example-files}
example_directory <- system.file("extdata", package = "fruitArchitecture")
list.files(example_directory)
```

# Recommended DEG-table workflow

```{r deg-workflow}
deg_table <- utils::read.csv(
  file.path(example_directory, "fruitArchitecture_example_deg.csv"),
  check.names = FALSE
)

annotation <- utils::read.csv(
  file.path(example_directory, "fruitArchitecture_example_annotation.csv"),
  check.names = FALSE
)

result <- fruitArchitectureFromDEG(
  deg_table = deg_table,
  annotation = annotation,
  species = "synthetic fruit example",
  n_permutations = 1000,
  seed = 1234
)

result
```

```{r result-tables}
result$module_summary
result$interface_summary
result$level3b_genes
```

# Figures

```{r module-figure, fig.width=8, fig.height=5, , fig.alt="Horizontal bar chart summarizing significant gene support across the default fruit architecture modules."}
print(
  plot(
    result,
    type = "modules"
  )
)
```

```{r interface-figure, fig.width=8, fig.height=5, , fig.alt="Horizontal bar chart summarizing strict gene overlap for each pairwise PHMIES interface."}
print(
  plot(
    result,
    type = "interfaces"
  )
)
```

```{r null-figure, fig.width=8, fig.height=5, , fig.alt="Permutation null distribution for the number of Level 3B genes, with the observed value shown as a dashed vertical line."}
print(
  plot(
    result,
    type = "null"
  )
)
```

# Export a complete example analysis

```{r export, eval=FALSE}
output_directory <- file.path(getwd(), "fruitArchitecture_example_output")

exportArchitecture(
  result,
  directory = output_directory,
  prefix = "synthetic_example"
)

exportFigures(
  result,
  directory = file.path(output_directory, "figures"),
  formats = c("pdf", "png", "tiff"),
  width = 9,
  height = 6,
  dpi = 600
)
```

# Raw-count workflow

```{r count-workflow, eval=FALSE}
counts_data <- utils::read.csv(
  file.path(example_directory, "fruitArchitecture_example_counts.csv"),
  row.names = 1,
  check.names = FALSE
)
counts <- as.matrix(counts_data)
storage.mode(counts) <- "integer"

design <- utils::read.csv(
  file.path(example_directory, "fruitArchitecture_example_design.csv"),
  row.names = 1,
  check.names = FALSE
)
design$condition <- factor(design$condition, levels = c("immature", "ripe"))

count_result <- fruitArchitecture(
  counts = counts,
  design = design,
  species = "synthetic fruit example",
  formula = ~ condition,
  contrast = c("condition", "ripe", "immature"),
  annotation = annotation,
  n_permutations = 1000,
  seed = 1234
)

count_result
```

# Input requirements

The DEG table must contain unique gene identifiers, log2 fold changes, and
adjusted P-values. Annotation must be long format, with one row for each
gene-module assignment. A strict Level 3B member is a significant gene assigned
to all three PHMIES core modules in the selected architecture definition.
